A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-29 paper/writeup.tex
Tweaks to ignorance and omission section

We began this study with the observation that the simple transformation
of \figref{fig:extending_diagram} would reduce the number of edges
in the succinct tree sequence representation of the ARG.
This is essentially a recombination-based parsimony argument,
and we have shown that this line of reasoning
leads to ARGs that are substantially more compact and faster to operate on,
and that contain more complete information about true ancestral relationships.
These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
Although a number of ARG inference methods may be taking advantage of this information,
it is our impression that this source of information is not widely appreciated.
In fact, due to the field's focus on marginal trees rather than haplotypes,
we had to develop a haplotype-aware measure of (dis)agreement between ARGs in order to study the accuracy of the proposed algorithm.

@@

\paragraph{Ignorance and omission in the ARG}
As motivation, we presented above a ``historical'' view of the ARG --
i.e., that each aspect of an inferred ARG is intended to represent
a portion of some particular historical genome
(for instance, the MRCA of some set of sampled genomes).
Furthermore, \figref[A\&B]{fig:conceptual} implicitly takes the position
that relationships \emph{not} depicted in the ARG are implied to not exist.
An alternative interpretation
of the ARG depicted in \figref[A\&B]{fig:conceptual} would be that we have no information as to 
how node 2 inherited from node 4 on the right-hand interval,
rather than saying that the line of transmission specifically did not pass
through node 3.
The ``simplification'' algorithm of \citet{kelleher2018efficient},
or the Hudson algorithm for coalescent simulation
as implemented in \msprime{} \citep{kelleher2016efficient},
each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;
so for ARGs produced by these algorithms, the correct interpretation is that
the omission of unary spans reflects a lack of knowledge.
In this paper, we have shown that, for the most part, this missing information can be imputed.

>>>>

We began this study with the observation that the simple transformation
of \figref{fig:extending_diagram} would reduce the number of edges
in the succinct tree sequence representation ARGs.
This is essentially a recombination-based parsimony argument,
and we have shown that this line of reasoning
leads to ARGs that are substantially more compact and faster to operate on,
and that contain more complete information about true ancestral relationships.
These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
Although a number of ARG inference methods may be taking advantage of this information,
it is our impression that this source of information is not widely appreciated.
In fact, due to the field's focus on marginal trees rather than haplotypes,
we had to develop a haplotype-aware measure of (dis)agreement between ARGs in order to study the accuracy of the proposed algorithm.

@@

\paragraph{Ignorance and omission in an ARG}
As motivation, we presented above a ``historical'' view of ARGs --
i.e., that each aspect of an inferred ARG is intended to represent
a portion of some particular historical genome
(for instance, the MRCA of some set of sampled genomes).
Furthermore, \figref[A\&B]{fig:conceptual} implicitly takes the position
that relationships \emph{not} depicted in an ARG are implied to not exist.
An alternative interpretation
of the ARG depicted in \figref[A\&B]{fig:conceptual} would be that we have no information as to 
how node 2 inherited from node 4 on the right-hand interval,
rather than saying that the line of transmission specifically did not pass
through node 3.
The ``simplification'' algorithm~\citep{kelleher2018efficient}
and the Hudson algorithm for coalescent
simulation~\citep{hudson1983properties,kelleher2016efficient}
each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;
so for ARGs produced by these algorithms, the correct interpretation is that
the omission of unary spans reflects a lack of knowledge.
In this paper, we have shown that, for the most part, this missing information can be imputed.
